---
title: "tradeseq_plots"
author: "Kin Lau"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    self_contained: yes
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
---

```{r keep_figures, cache=TRUE}
# this chunk is just to keep the _files directory even when we turn off cacheing
```

```{r starttime}
# save start time for script
start_tm <- Sys.time()
start_tm
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, cache.lazy = FALSE, dev=c('png','pdf'), fig.width=8, fig.height=8)

```

```{r make_outdir}
outdir <- "./tradeseq_plots_out_files/"

dir.create(outdir, recursive=TRUE)
```

# Packages loaded

```{r loadlibs, echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE}
library(dplyr)
library(stringr)
library(readr)
library(tradeSeq)
library(pheatmap)
library(ggplotify)
library(dittoSeq)
library(scran)
library(scater)
library(patchwork)
library(slingshot)
library(openxlsx)
library(ggplot2)
library(viridis)
```

# Read in results

```{r read_results}
sce_slingshot <- readRDS("../../pseudotime_out_files/slingshot_sce.rds")
sce <- readRDS("../tradeseq_out_files/tradeseq_sce.rds")
ATres <- readRDS("../tradeseq_out_files/ATres.rds")

```

# Replot pseudotime trajectories

```{r plot_trajectories, fig.width=14, fig.height=4}
pseudotime_cols <- grep(colnames(colData(sce_slingshot)), pattern = "slingPseudo", value = TRUE)
names(pseudotime_cols) <- pseudotime_cols
gg_list <- lapply(pseudotime_cols, function(x) {
  #plotUMAP(sce_slingshot, colour_by=x) + 
  dittoDimPlot(sce_slingshot, x) +
    ggtitle(str_replace(x, "slingPseudotime", "Lineage"))
}) 

slingcurves <- slingCurves(sce_slingshot)

# draw all lineages in each panel
gg_list_w_curves <- gg_list
lin_colors <- viridis(length(gg_list_w_curves))
for (i in 1:length(slingcurves)) {
  path <- slingcurves[[i]]
  trajectory <- data.frame(path$s[path$ord,])
  gg_list_w_curves <- lapply(gg_list_w_curves, function(x) x + geom_path(data=trajectory, aes(x=UMAP_1, y=UMAP_2), size=1.2, color=lin_colors[i]))
  #gg <- gg + geom_path(data=trajectory, aes(x=UMAP_1, y=UMAP_2), size=1.2, color=i)
}
wrap_plots(gg_list_w_curves)

# draw single lineage in each panel
wrap_plots(
  lapply(1:length(slingcurves), function(x){
    path <- slingcurves[[x]]
    trajectory <- data.frame(path$s[path$ord,])
    gg_list[[x]] + geom_path(data=trajectory, aes(x=UMAP_1, y=UMAP_2), size=1.2, color=lin_colors[x])
  })
)
```

# Replot pseudotime trajectories with clusters

```{r plot_trajectories_w_clusters, fig.width=7, fig.height=5}
cluster_res <- "integrated_snn_res.0.3"

umap_w_slings <- dittoDimPlot(sce_slingshot, cluster_res)

slingcurves <- slingCurves(sce_slingshot)

for (i in 1:length(slingcurves)) {
  path <- slingcurves[[i]]
  trajectory <- data.frame(path$s[path$ord,])
  umap_w_slings <- umap_w_slings + geom_path(data=trajectory, aes(x=UMAP_1, y=UMAP_2), size=1.2, color=lin_colors[i])
}
umap_w_slings

```

# Plot sample proportion in each lineage

```{r lineage_prop, fig.width=10, fig.height=4}
pseudotime_cols <- grep(colnames(colData(sce)), pattern = "slingPseudo", value=TRUE)

for (i in 1:length(pseudotime_cols)){
  sce[[paste0("memb_of_lin_", i)]] <- ifelse(!is.na(colData(sce)[[paste0("slingPseudotime_", i)]]), "Member", "NotMember")
}

wrap_plots(
  lapply(1:length(pseudotime_cols), function(x){
    dittoBarPlot(sce, paste0("memb_of_lin_", x), group.by = "Sample", 
                 split.by = "Tissue", retain.factor.levels=TRUE, 
                 split.adjust=list(scales="free_x"),
                 main=paste0("Lineage ", x))
  })
) + plot_layout(guides="collect")
```

# Plot sample proportion in each lineage, combined

```{r lineage_prop_combined, fig.width=5, fig.height=4}
lin_assignments <- slingBranchID(sce_slingshot)
stopifnot(identical(names(lin_assignments), colnames(sce)))
sce$lin_assignments <- lin_assignments
dittoBarPlot(sce, "lin_assignments", group.by = "Sample", 
             split.by = "Tissue", retain.factor.levels=TRUE, 
             split.adjust=list(scales="free_x"),
             main = "Lineage assignments")
```


# Plot heatmaps of top genes associated with pseudotime along each inferred lineage

```{r pseudotime_genes_heatmap, fig.width=8, fig.height=14}
lineage_pvals <- grep("^pvalue_\\d+$", colnames(ATres), value=TRUE)
lineage_wald <- grep("^waldStat_\\d+$", colnames(ATres), value=TRUE)

ht_plots <- lapply(1:length(lineage_pvals), function(x){
  lin_genes <-  ATres[which(p.adjust(ATres[[paste0("pvalue_", x)]], "fdr") < 0.05), ]
  lin_genes <- lin_genes[order(lin_genes[[paste0("waldStat_", x)]], decreasing = TRUE), ]
  yhatSmooth <- predictSmooth(sce, gene = rownames(lin_genes)[1:max(c(length(lin_genes), 100))], 
                              nPoints = 50, tidy = FALSE)
  lin_name <- paste0("lineage", x)
  as.ggplot(
    pheatmap(t(scale(t(yhatSmooth[, str_detect(colnames(yhatSmooth), paste0(lin_name, "_"))]))),
             cluster_cols = FALSE,
             show_rownames = TRUE, 
             show_colnames = FALSE, 
             main=lin_name, 
             silent=TRUE)
  )
})
ht_plots
#wrap_plots(ht_plots)

```



# Output results of associating each gene to pseudotime lineages


```{r out_pseduotime_genes}
ATres_cols <- colnames(ATres)
ATres_cols <- ATres_cols[-length(ATres_cols)]

ATres_cols_split <- split(ATres_cols, rep(1:(length(ATres_cols)/3), each = 3))
names(ATres_cols_split) <- c("global", paste0("lineage_", 1:(length(ATres_cols_split)-1)))

ATres_split <- lapply(ATres_cols_split, function(x) {
  stopifnot(all(!is.na(rownames(ATres))))
  
  ATres[, x] %>% #c(x, "meanLogFC")] %>%
    tibble::rownames_to_column("gene") %>%
    dplyr::rename_all(., function(y) str_remove(y, "_\\d+$")) %>%
    dplyr::mutate(FDR=p.adjust(pvalue, "fdr")) %>%
    dplyr::arrange(desc(waldStat)) %>% # pvalue) %>%
    dplyr::filter(apply(., 1, function(y) sum(is.na(y))) < length(x)) # remove untested or unsuccessfully tested genes
    #dplyr::filter(FDR < 0.05) # filter for significant genes only
})

write.xlsx(ATres_split, paste0(outdir, "pseudotime_genes.xlsx"), overwrite = TRUE)
```

# Pete's genes

## Plot counts along pseudotime

```{r plotSmoothers_prep_pete}
genes <- c("S100a8", "S100a9", "S100a11", "Bcl2", "Bcl2l11", "Zeb2", "Vav3", "Cd93", "Ctla4")
smoother_plots <- lapply(genes, function(x) plotSmoothers(sce, counts(sce), gene = x) + ggtitle(x))
```

```{r plotSmoothers_pete, fig.height=10, fig.width=12}
wrap_plots(smoother_plots)
```

## Ly6c2 smoother plot

```{r Ly6c2_smoother, fig.width=6.5, fig.height=5}
smoother_plot_gene <- "Ly6c2"
plotSmoothers(sce, counts(sce), gene = smoother_plot_gene) + ggtitle(smoother_plot_gene)
```

```{r umap_gene_counts_pete, fig.width=10, eval=FALSE}

# multi_dittoDimPlot(sce_slingshot, genes, assay="logcounts",
#                    order="increasing", size=0.7,
#                    legend.show=TRUE, 
#                    add.trajectory.curves = slingCurves(sce_slingshot))
```


# Top 12 Lineage 1

## Plot counts along pseudotime

```{r plotSmoothers_prep_topLin1}
genes <- ATres_split$lineage_1$gene[1:12]
smoother_plots <- lapply(genes, function(x) plotSmoothers(sce, counts(sce), gene = x) + ggtitle(x))
```

```{r plotSmoothers_topLin1, fig.height=10, fig.width=16}
wrap_plots(smoother_plots, nrow = 3)
```


# Top 12 Lineage 2

## Plot counts along pseudotime

```{r plotSmoothers_prep_topLin2}
genes <- ATres_split$lineage_2$gene[1:12]
smoother_plots <- lapply(genes, function(x) plotSmoothers(sce, counts(sce), gene = x) + ggtitle(x))
```

```{r plotSmoothers_topLin2, fig.height=10, fig.width=16}
wrap_plots(smoother_plots, nrow = 3)
```


# Top 12 Lineage 3

## Plot counts along pseudotime

```{r plotSmoothers_prep_topLin3}
genes <- ATres_split$lineage_3$gene[1:12]
smoother_plots <- lapply(genes, function(x) plotSmoothers(sce, counts(sce), gene = x) + ggtitle(x))
```

```{r plotSmoothers_topLin3, fig.height=10, fig.width=16}
wrap_plots(smoother_plots, nrow = 3)
```

# SessionInfo

```{r sessioninfo}
sessionInfo()
```

# Time

```{r endtime}
# output time taken to run script
end_tm <- Sys.time()
end_tm
end_tm - start_tm

```
